Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2021-12-02 Manuscript/Manuscript.Rmd
Full text version of introduction and first part case study 1

```{r patterns, echo=FALSE}
# missingness indicator plot (monotone vs non-monotone)
miss_ind <- tibble(
    rownr = 5:1,
    unit = rep("observed", 5),
    cluster = rep("observed", 5),
    X1 = rep("observed", 5),
    X2 = c(rep("observed", 3), rep("missing", 2)),
    X3 = c("missing", rep("observed", 3), "missing")
  ) %>% pivot_longer(cols = c(unit, cluster, X1, X2, X3))

@@

### The incomplete data

- What does the missingness look like? Induced MAR/MNAR missingness. Missing data pattern:

```{r pop_pat, echo=FALSE, fig.height=6, fig.width=6, message=FALSE, warning=FALSE}
# missingness
plot_md_pat(pop)
# TODO: add pattern by cluster, text mining example square size
# TODO: try to make it a shepley plot 
```

- What are the ICCs? For `popular` the ICC is `r round(icc_pop, 2)`. For `popteach` it is `r round(icc_teach, 2)`. It would be wise to use multilevel modeling.

- Does the missing data of `popular` depend on `popteach`? Does the missingness in teacher popularity depend on pupil popularity? -> Check this by making a histogram of `popteach` separately for the pupils with known popularity and missing popularity, and the other way around.

```{r pop_dist, echo=FALSE, message=FALSE, warning=FALSE}
# distributions
# plot_NA_cond(pop, x = "popteach", z = "popular", bins = 1)
# plot_NA_cond(pop, x = "popular", z = "popteach", bins = 1)
# TODO: check if up-side-down plot works
# or maybe overlapping densities
# TODO: check smoothing in geom_density function and make it the inverse of the sample size
# TODO: add functions to mice

@@

-  We do see that the distribution for the missing `popular` is further to the right than the distribution for observed `popular`. This would indicate a right-tailed MAR missingness. In fact this is exactly what happens, because we created the missingness in these data ourselves. But we made it observable by examining the relations between the missingness in popular and the observed data in `popteach`. There is also a dependency between the missingness in teacher popularity and pupil popularity. The relation seems to be right-tailed as well.

### The imputation model

- We could ignore the multilevel structure of the data. This is likely invalid, given the high ICCs, but we'll do it anyways.

- We'll use predictive mean matching to impute the continuous variables (some appear to be somewhat ordinal), and logistic regression to impute the binary variable `sex`. We do not use the observation identifier `pupil` or cluster identifier `class` as predictors to impute other variables.

```{r pop-ignored}
# dry run to get imputation parameters
ini <- mice(pop, maxit = 0)

# extract predictor matrix and adjust
pred <- ini$pred
pred[, c("class", "pupil")] <- 0
pred

# impute the data, ignoring the cluster structure
imp_ignored <- mice(pop, maxit = 10, pred = pred, print = FALSE)

# check convergence of the imputation model
plot(imp_ignored)

# compare descriptives before and after imputation
psych::describe(pop)[, c("n", "mean", "median", "min", "max", "sd")]
psych::describe(mice::complete(imp_ignored))[, c("n", "mean", "median", "min", "max", "sd")] #note that this is just 1 imputation, not the pooled results
# TODO: add stripplot with boxplot overlay instead of the tables (make pooled one thick on top)
# TODO: pool mean median and sd
# feedback Stef: numbers, continuous statistics such as means, and uncertainty estimates. So we can pool the sd's. And leave out the min and max, because those are not normally distr.
# TODO: add FMI for each of the estimates? at least for the mean

@@

# compare ICCs before and after imputation
ICCs <- data.frame(
  vars = c("popular", "popteach", "texp"), 
  incomplete = c(multilevel::ICC1(aov(popular ~ class, pop)), 
               multilevel::ICC1(aov(popteach ~ class, pop)),
               multilevel::ICC1(aov(texp ~ class, pop))), 
  ignored = c(multilevel::ICC1(aov(popular ~ class, complete(imp_ignored))), 
              multilevel::ICC1(aov(popteach ~ class, complete(imp_ignored))), 
              multilevel::ICC1(aov(texp ~ class, complete(imp_ignored))))
  )
ICCs

- As the original ICCs show, 100% of the variance in `texp` can be attributed to the clustering variable `class`. This tells us that the multilevel structure of the data should be taken into account. If we don't, we'll end up with incorrect imputations, biasing the effect of the clusters towards zero.

- We can also observe that the teacher experience increases slightly after imputation. This is due to the MNAR missingness in `texp`. Higher values for `texp` have a larger probability to be missing. This may not a problem, however, if at least one pupil in each class has teacher experience recorded, we can deductively impute the correct (i.e. true) value for every pupil in the class.

- We'll now use `class` as a predictor to impute all other variables.

```{r pop-predictor}
# adjust the predictor matrix
pred <- ini$pred 
pred[, "pupil"] <- 0
pred

# impute the data, cluster as predictor
imp_predictor <- mice(pop, maxit = 10, pred = pred, print = FALSE)

# check logged events
head(imp_predictor$loggedEvents)
## "The mice() function detects multicollinearity, and solves the problem by removing one or more predictors for the model", in this case texp is removed as predictor of popular and popteach.

# check convergence of the imputation model
plot(imp_predictor)

# compare descriptives before and after imputation
psych::describe(pop)[, c("n", "mean", "median", "min", "max", "sd")]
psych::describe(mice::complete(imp_predictor))[, c("n", "mean", "median", "min", "max", "sd")] #note that this is just 1 imputation, not the pooled results

@@

- Now, we can clearly see that the imputed values of `texp` are higher than the observed values, which is in line with right-tailed MNAR.

- The ICCs are way more in line with the ICCs in the incomplete data. But this is a quick and dirty way of imputing multilevel data. We *should* be using a multilevel model.

@@

```{r impact, echo=FALSE}
# load data
data("impact")
# # descriptive statistics
# by(impact, impact$name, summary) 
# psych::describe(impact)[,c(2:5,8:9)]
# missingness
plot_md_pat(impact)
# TODO: rotate variable labels

>>>>

```{r patterns, fig.height=4, fig.width=3.5, fig.cap = "Missingness in multilevel data", echo=FALSE}
# missingness indicator plot (monotone vs non-monotone)
miss_ind <- tibble(
    rownr = 5:1,
    unit = rep("observed", 5),
    cluster = rep("observed", 5),
    X1 = rep("observed", 5),
    X2 = c(rep("observed", 3), rep("missing", 2)),
    X3 = c("missing", rep("observed", 3), "missing")
  ) %>% pivot_longer(cols = c(unit, cluster, X1, X2, X3))

@@

### Incomplete data

```{r pop_pat, echo=FALSE, fig.height=6, fig.width=4, fig.cap = "Missing data pattern in the popularity data", message=FALSE, warning=FALSE}
# missingness
plot_md_pat(pop)
# TODO: add pattern by cluster, text mining example square size
# TODO: try to make it a shepley plot 
```

The popularity data is created such that there are strong relations between the incomplete variables and the clustering variable `class`. We can express this using the intra-class correlation (ICC). For `popular` the ICC is `r round(icc_pop, 2)`. For `popteach` it is `r round(icc_teach, 2)`. It would thus be wise to use multilevel modeling.

The missingness in this dataset is induced conform MAR and MNAR mechanisms. The missing data pattern, Figure \ref{fig:pop_pat}, shows the systematic nature of the missingness.

To develop the best imputation model, we need to know whether the missingness in one variable depends on the observed values of other variables. Visual inspection usually suffices. We'll highlight only two variables to illustrate, but ideally one would inspect all relations. The questions we'll ask are: 'Does the missing data of `popular` depend on `popteach`?' and 'Does the missingness in teacher popularity depend on pupil popularity?' We'll evaluate this by making a histogram of `popteach` separately for the pupils with known popularity and missing popularity, and the other way around.

In Figure \ref{fig:pop_dist} we see that the distribution for the missing `popular` is further to the right than the distribution for observed `popular`. This would indicate a right-tailed MAR missingness. In fact, this is exactly what happens, because the missingness in these data was created manually. Now, we've made it observable by examining the relations between the missingness in popular and the observed data in `popteach`. There is also a dependency between the missingness in teacher popularity and pupil popularity. The relation seems to be right-tailed as well.

```{r pop_dist, fig.cap = "Conditional distributions in the popularity data", echo=FALSE, message=FALSE, warning=FALSE}
# distributions
# plot_NA_cond(pop, x = "popteach", z = "popular", bins = 1)
# plot_NA_cond(pop, x = "popular", z = "popteach", bins = 1)
# TODO: check if up-side-down plot works
# or maybe overlapping densities
# TODO: check smoothing in geom_density function and make it the inverse of the sample size
# TODO: add functions to mice

@@

### Imputation model

The first imputation model that we'll use is likely to be invalid. In this model, we ignore the multilevel structure of the data, despite the high ICCs. This is purely to illustrate the effects of ignoring the clustering in our imputation effort.

We'll use predictive mean matching to impute the continuous variables and logistic regression to impute the binary variable `sex`. We do not use the observation identifier `pupil` or cluster identifier `class` as predictors to impute other variables.

```{r pop_ignored, echo=TRUE, message=FALSE, warning=FALSE}
# dry run to get imputation parameters
ini <- mice(pop, maxit = 0)

# extract predictor matrix and adjust
pred <- ini$pred
pred[, c("class", "pupil")] <- 0

# impute the data, ignoring the cluster structure
imp_ignored <- mice(pop, maxit = 10, pred = pred, print = FALSE)
```

### Imputed data

```{r pop_ignored_eval, echo=FALSE, message=FALSE, warning=FALSE}
# # check convergence of the imputation model
# plot(imp_ignored)

# # compare descriptives before and after imputation
# psych::describe(pop)[, c("n", "mean", "median", "min", "max", "sd")]
# psych::describe(mice::complete(imp_ignored))[, c("n", "mean", "median", "min", "max", "sd")] #note that this is just 1 imputation, not the pooled results
# TODO: add stripplot with boxplot overlay instead of the tables (make pooled one thick on top)
# TODO: pool mean median and sd
# feedback Stef: numbers, continuous statistics such as means, and uncertainty estimates. So we can pool the sd's. And leave out the min and max, because those are not normally distr.
# TODO: add FMI for each of the estimates? at least for the mean

@@

# compare ICCs before and after imputation
ICCs <- data.frame(
  vars = c("popular", "popteach", "texp"), 
  incomplete = c(multilevel::ICC1(aov(popular ~ class, pop)), 
               multilevel::ICC1(aov(popteach ~ class, pop)),
               multilevel::ICC1(aov(texp ~ class, pop))), 
  ignored = c(multilevel::ICC1(aov(popular ~ class, complete(imp_ignored))), 
              multilevel::ICC1(aov(popteach ~ class, complete(imp_ignored))), 
              multilevel::ICC1(aov(texp ~ class, complete(imp_ignored))))
  )
ICCs
```

As the original ICCs show, 100% of the variance in `texp` can be attributed to the clustering variable `class`. This tells us that the multilevel structure of the data should be taken into account. If we don't, we'll end up with incorrect imputations, biasing the effect of the clusters towards zero.

We can also observe that the teacher experience increases slightly after imputation. This is due to the MNAR missingness in `texp`. Higher values for `texp` have a larger probability to be missing. This may not a problem, however, if at least one pupil in each class has teacher experience recorded, we can deductively impute the correct (i.e. true) value for every pupil in the class.

### Imputation model

We'll now use `class` as a predictor to impute all other variables. This is still not recommended practice, since it only works under certain circumstances and results may be biased. But at least, it includes some multilevel aspect. Colloquially, this is 'multilevel imputation for dummies'.

```{r pop_predictor, echo=FALSE, message=FALSE, warning=FALSE}
# adjust the predictor matrix
pred <- ini$pred 
pred[, "pupil"] <- 0

# impute the data, cluster as predictor
imp_predictor <- mice(pop, maxit = 10, pred = pred, print = FALSE)
```

### Imputed data

```{r pop_predictor_eval, echo=FALSE, message=FALSE, warning=FALSE}
# # check logged events
# head(imp_predictor$loggedEvents)
## "The mice() function detects multicollinearity, and solves the problem by removing one or more predictors for the model", in this case texp is removed as predictor of popular and popteach.

# # check convergence of the imputation model
# plot(imp_predictor)

# # compare descriptives before and after imputation
# psych::describe(pop)[, c("n", "mean", "median", "min", "max", "sd")]
# psych::describe(mice::complete(imp_predictor))[, c("n", "mean", "median", "min", "max", "sd")] #note that this is just 1 imputation, not the pooled results

@@

Now, we can clearly see that the imputed values of `texp` are higher than the observed values, which is in line with right-tailed MNAR.

The ICCs are way more in line with the ICCs in the incomplete data. But this is a quick and dirty way of imputing multilevel data. We *should* be using a multilevel model.

### Imputation model

To include...

@@

```{r impact, echo=FALSE}
# load data
data("impact")
# # descriptive statistics
# by(impact, impact$name, summary) 
# psych::describe(impact)[,c(2:5,8:9)]
# # missingness
# plot_md_pat(impact)
# TODO: rotate variable labels
